---
title: "Regression Analysis"
author: "Pier Lorenzo Paracchini"
date: "January 4, 2016"
output: 
  html_document: 
    keep_md: yes
---
```{r echo = FALSE, results='hide'}
Sys.setlocale("LC_ALL", "C")
```

Regression analysis is the preferred method when both the response variable and the predictors (explanatory variables) are __continuous variables__.

The essence of the Regression Analysis is to use sample data to estimate parameter values and their standard errors having selected a model which describes the relationship between the response and the predictors (explanatory variables). __Note!__ _The model choice is a really big deal._

The __simplest model__ is the __linear model__

$y = a + b * x$

where

* y: response variable
* x: explanatory variable
* a: the __intercept__, the value of y when x = 0
* b: the __slope__, the change of y for a change of x

##Example
```{r loadingExampleData}
reg.data <- read.csv("./datasets/tannin.csv")
attach(reg.data)
```

```{r summaryData}
str(reg.data)
summary(reg.data)

#Regression by eye: 
#What has happened to the y value?
#How has the x value changed?
plot(tannin, growth, pch=21, bg="blue")
```

###Linear Regression
```{r linearRegressionModel}
#growth = a + b * tannin
reg.model.fit <- lm(growth ~ tannin)
#Calculate the predicted y based on the model (for the provided x)
reg.model.prediction <- predict(reg.model.fit)

summary(reg.model.fit)
```

The two parameters __a__ and __b__ are called __coefficients__ in R and their estimates are calculated using the __maximum likelihood estimates__ of the parameters. Alias _"we want to find the values of the slope and intercept that make the data most likely."_ - implemented using SSE (Sum of Squared Errors) and minimizing it.

```{r visualizingFitModel}
plot(tannin, growth, pch=21, bg="blue")
abline(reg = reg.model.fit, col="green")
```

The fit is reasonably good but it is not perfect. The observations do not lie on the fitted line. The difference between an observation and the predicted value (using the model) for a specific x (explanatory variable) is called __residual__.

The fit pass through the point in the centre of the cloud of data. The best-fit line will be pivoted about the mean values of x and y and our job is to find the best value for this slope – the one that minimizes the sum of squares of the red lines in the graph above.

```{r predictionAndResidualsFitModel, collapse=TRUE}
#Observations: growth (y)
growth

#Predictions
reg.model.prediction

#Residual (Observation - Prediction)
growth - reg.model.prediction
```

Let's visualize the __residuals__ ...

```{r visualizingFitModel_extended}
plot(tannin, growth, pch=21, bg="blue")
abline(reg = reg.model.fit, col="green")
for(i in 1:length(tannin)){
  lines(c(tannin[i], tannin[i]), c(growth[i], reg.model.prediction[i]), col="red")
}
```

Residuals:

$d_i = y_i - \hat y_i$  
$d_i = y_i - \hat a - \hat b * x_i$ 

__Property#1__: The sum of the residuals in a linear regression is __zero__ $\sum d_i = \sum (y_i - \hat a - \hat b * x_i) = 0$   


Residuals are very important to estimate and evalute the goodness of the model. Our __maximum likelihood__ model is defined as _the model that minimize the sum of the squares of these residuals_.

Visualizing the SSE changing the slope coefficient. We can see that the minimum SSE is at the estimated slope coeffiecient (using the linear regression model).

```{r sseGraph, collapse = TRUE}
b <- seq(-3, 1, 0.01)
sse <- numeric(length = length(b))

for (i in 1:length(b)){
  #Based on the fact that the regression line pass through the sample mean of x and y
  a <- mean(growth) - b[i] * mean(tannin)
  evaluated.residuals <- growth - (a + (b[i] * tannin))
  sse[i] <- sum(evaluated.residuals ^ 2)
}

#estimated b for the minimum SSE
b[which(sse == min(sse))]

plot(b, sse, type="l", ylim=c(19,25))
abline(h=20.07225,col="green",lty=2)
arrows(-1.216,20.07225,-1.216,19,col="red")
```

###Calculation Involved in Linear Regression

Basic elements: the __corrected sums__ calculations.l

Corrected Sum of Squares of y: $SSY = \sum y_i^2 - (\sum y_i)^2 / n$  
Corrected Sum of Squares of x: $SSX = \sum x_i^2 - (\sum x_i)^2 / n$  
Corrected Sum of Squares of products xy: $SSXY = \sum x_i * y_i - (\sum x_i * \sum y_i)  / n$  

These corrected sum are absolutely central to everything that follows about regression and analysis of variance. Specifically ...

Explained variation is the __Regression Sum of Squares__: $SSR = SSXY^2/SSX$  
Unexplained variation is the __Error Sum of Squares__: $SSE = SSY - SSR = \sum (y_i - \hat a - \hat b * x_i)^2$  
Correlation Coefficient __r__: $r = SSXY / (\sqrt{SSX * SSY})$  

The maximum likelihood estimate for the slope b and the intercept a can be calculated using the folowing formulas:

$b = SSXY/SSX$  
$a = \overline{y} - b * \overline{x}$, where $\overline{y} = \sum y_i / n$ and $\overline{x} = \sum x_i / n$  


Estimation of the coefficients done using the previous formulas are identical to estimation done using the `lm` function.

```{r correctedSumAndCoefficients, collapse=TRUE}
n <- length(growth)
SSY <- sum(growth ^ 2) - sum(growth) ^ 2 / n
SSX <- sum(tannin ^ 2) - sum(tannin) ^ 2 / n
SSXY <- sum(growth * tannin) - (sum(growth) * sum(tannin))/n

b_coeff <- SSXY/SSX
a_coeff <- mean(growth) - b_coeff * mean(tannin)
#Intercept
a_coeff
#Slope
b_coeff
```

In addition to the estimates of the parameters we need to measure the __unreliability associated with each one of them__. In other words we need to calculate the __standard error of the intercept__ and the __standard error of the slope__.

#Reference

* Chapter 7 of ["Statistics: An Introduction Using R"](http://eu.wiley.com/WileyCDA/WileyTitle/productCd-1118941098.html)