---
title: "Caret Package"
author: "Pier Lorenzo Paracchini"
date: "February 9, 2016"
output: 
  html_document: 
    keep_md: yes
---

The `caret` package is a set of functions that attemt to streamline the process of creating predictive models. It includes functions to:

* preprocess the data
* split the data
* train models (fit) and test nodels
* comparison of models

`caret` acts as a **wrapper** around other packages used for predictive models.

##Packages, Installations (with Dependencies)

```{r eval = FALSE}
#The caret package
install.packages("caret", dependencies = c("Depends", "Suggests"))

#Binning Support
install.packages("Hmisc")

#Multiple plots in same grid
install.packages("gridExtra")

#Data
install.packages("ISLR")
install.packages("AppliedPredictiveModeling")
```

```{r loadingRequiredPackages, echo = FALSE, message=FALSE, warning=FALSE}
require(caret)
require(Hmisc)
require(gridExtra)
require(ggplot2)
```


##The datasets  
The `ISLR::Wage` dataset:  
```{r theData1, message=FALSE, warning=FALSE, echo = FALSE}
require(ISLR)
data(Wage)
str(Wage)
summary(Wage)
```

The `iris` dataset:  
```{r theData2, message=FALSE, warning=FALSE, echo = FALSE}
data(iris)
str(iris)
summary(iris)
```

#The Caret Vocabulary

##Plotting the Predictors::Visualizations
**Important!** When performing exploratory analysis on the available dataset is a **good preactice** to **split the available data set in a training dataset and a testing dataset** and perform the relevant exploration on the **training dataset**.

Visual inspection/ exploration is important for building up an "initial" understanding of the available predictors, "possible" patterns and relationships. E.g. looking for skewness, outliers, ...

##Plots
###Scatterplot Matrix (caret)
The `pairs` plot option is available for **regression** and **classification** problems.

```{r visualizationData1}
featurePlot(x=Wage[, c("age", "education", "jobclass")], y=Wage$wage, plot="pairs")
```

```{r visualizationData2}
featurePlot(x=iris[, 1:4], y=iris$Species, plot="pairs", auto.key = list(colums= 3))
```

###Scatterplot Matrix with Ellipses (caret)
The `ellipse` plot option is available for **classification** problems.

```{r visualizationData2_1}
featurePlot(x=iris[, 1:4], y=iris$Species, plot="ellipse", auto.key = list(colums= 3))
```

###Scatterplot (ggplot2)
Simple Scatterplot:  
```{r}
qplot(x=age, y=wage, data=Wage)
```

Adding the `jobclass` dimension:  
```{r}
qplot(x=age, y=wage, colour=jobclass, data=Wage)
```

Using `education` dimension and adding regression smoothers:  
```{r}
qq <- qplot(x=age, y=wage, colour=education, data=Wage)
qq + geom_smooth(method="lm", formula = y~x)
```

###Boxplot (Hmisc + ggplot2 + gridExtra)
A boxplot example using `qplot` ...  
```{r}
cutWage <- cut2(Wage$wage, g=3) #g: number of quantiles group
p1 <- qplot(x = cutWage, y = age , data=Wage, fill=cutWage, geom=c("boxplot"))
p1
```

Adding the points overlayed...  
```{r}
p2 <- qplot(x = cutWage, y = age , data=Wage, fill=cutWage, geom=c("boxplot", "jitter"))
grid.arrange(p1, p2, ncol=2)
```

###Boxplot (caret)
A boxplot example using `featurePlot` ...  
```{r}
featurePlot(x = iris[, 1:4],
            y = iris$Species,
            plot = "box",
            ##pass in options to bwplot
            scales = list(y = list(relation="free"), x = list(rot = 90)),
            layout= c(4,1), auto.key = list(columns = 2))
```

###Tables
```{r}
t1 <- table(cut2(Wage$wage, g=3), Wage$jobclass)
##Show table content
t1

##Show Proportion by row
prop.table(t1, 1)

##Show Proportion by col
prop.table(t1, 2)
```

##Pre-processing
###Binning: making factors out of quantitative predictors (Hmisc)
An example on how to break a quantitative variables into different categories.

```{r binningExample}
cutWage <- cut2(Wage$wage, g=3) #g: number of quantiles group
table(cutWage)
```