---
title: "Data Visualization"
author: "ppar"
output: html_document
---

Examples based on the following books

* ["R Data Visualization Cookbook"]([https://www.packtpub.com/packtlib/book/Big%20Data%20&%20Business%20Intelligence/9781783989508), Atmajitsinh Gohil.


#Visualization1
##Basic Scatterplot
```{r data1}
#install.packages("ISLR",  repos="http://cran.rstudio.com/")
library(ISLR)
attach(Carseats)
head(Carseats)
```

```{r plot}
#Creating a basic scatterplot
plot(Income, Sales, col = c(Urban), pch = 20, main = "Sales of Child Car Seats", xlab = "Income (000's of dollars)", ylab = "Unit Sales (in 000's)")
#Adding a legend
legend("topright", cex = 0.6, fill = c("red", "black"), legend = c("yes", "no"))
```

`col` used to assign a color to the data points and in this specific case a qualitative data colum `Urban` is used to color the points.

`pch` defines the point symbol - see `?point` for more info.

`cex` character expansion factor.

#Visualization2
##Scatterplot with texts, lables and lines

```{r data2}
child <- read.csv("./data/chlmort.csv", header = TRUE, sep = ",")
```
```{r data2Info}
summary(child)
str(child)
head(child)
```
```{r plotData2}
plot(child$gdp_bil, child$child, pch = 20, col = "#756bb1", xlim = c(0,max(child$gdp_bil)), ylim = c(0, 190), xlab = "GDP in billions in current US$", ylab = "Child Mortality Rate", main = "Child Mortality Rate in selected countries for 2012")

#Plot horizontal or vertical lines arg h=() - for horizontal
abline(h = (36.18), lwd = 1, col = "blue")

#Add labels to extreme points
text(8000,25, labels = c("Luxemburg"), cex = 0.75)
text(700,182, labels = c("Sierra Leone"), col = "red", cex = 0.75)
text(4000,50, labels = c("Average Child Mortality"), col = "blue", cex = 0.75)
```
```{r plot2Data}
##Adding a trend line
scatter.smooth(child$gdp_bil, child$child, pch = 20, lwd =0.75, col = "Blue", lpars = list(lty = 3, col ="black", lwd = 2),xlab ="GDP in Billions in current US$", ylab ="Child Mortality rate", main = "child Mortality Rate in selected countries for 2012" )
#Note! lpars() function to beautify the trend line
```
