---
title: "Data Frame Management using dplyr"
author: "ppar"
date: "28 Oct 2015"
output:
  html_document:
    keep_md: yes
    toc: no
---
```{r echo = FALSE, results="hide"}
Sys.setlocale("LC_ALL", "C")
```
#A grammar for data manipulation
The **dplyr** package is an important tool for dealing and managing data frames. It does not provide any news functionality to R, everything that can be done with **dplyr** can be done with base R but **dplyr** does **greatly simplify** existing functionality in R.

The **dplyr** package provides a "vocabulary"/ "grammar" used for data manipulation and for operating on data frames. And **dplyr** functions are faster and more performant than base R.

##The **vocabulary**/ **grammar**: the basic elements

* [`select`](#sel)
* [`filter`](#fil) (`slice`)
* [`arrange`](#arr)
* [`rename`](#ren)
* [`mutate`](#mut)
* [`group_by`](#gro)
* [`summarize`](#gro)
* `%>%`

##Installing the **dplyr** package
```{r installation, eval = FALSE}
install.packages("dplyr")
```

##Loading the **dplyr** package & the data
```{r loadPackage, warning = FALSE, message = FALSE, collapse = TRUE}
library(dplyr)
```
```{r checkVersion, collapse = TRUE}
packageVersion("dplyr")
```

Loading the data used for the code chunks.
```{r loadData, collapse = TRUE, cache = TRUE}
rawData <- read.csv("./data/stormData.csv", stringsAsFactors = FALSE)
str(rawData)
```

Preparing the data - specifically extracting the `BGN_DATE` (begin date) in order to transform it to a `Date` format, and adding a `TMP_BGN_YEAR` variable.

```{r bngDateTransformation, collapse = TRUE}
#BGN_DATE seems to have the following format
#d(max 2)/d(max2)/d(4) 0:00:00

#Verify that it is a correct assumption -> assumption evaluation should be TRUE
all(grepl("^\\d{1,2}\\/\\d{1,2}\\/\\d{4}\\s0:00:00$", rawData$BGN_DATE))

#Create a new col (temp feature): TMP_BGN_DATE
rawData$TMP_BGN_DATE <- gsub("\\s0:00:00$", "", rawData$BGN_DATE)
head(rawData[,c("BGN_DATE", "TMP_BGN_DATE")])
tail(rawData[,c("BGN_DATE", "TMP_BGN_DATE")])

#Transform TMP_BGN_DATE (character) to (Date)
rawData$TMP_BGN_DATE <- as.Date(rawData$TMP_BGN_DATE, format = "%m/%d/%Y")
head(rawData[,c("BGN_DATE", "TMP_BGN_DATE")])
tail(rawData[,c("BGN_DATE", "TMP_BGN_DATE")])
summary(rawData$TMP_BGN_DATE)

#Adding a TMP_BGN_YEAR variable
rawData$TMP_BGN_YEAR <- as.POSIXlt(rawData$TMP_BGN_DATE)$year + 1900
summary(rawData$TMP_BGN_YEAR)
```

##<a id="sel">`select`verb</a>
The `select` function is used to select the columns of a data frame that you want to focus on.

###Select cols by col numbers  
```{r selectExample1, collapse=TRUE}
#Select only the following columns 
#STATE__(1), BGN_DATE(2), BGN_TIME(3), TIME_ZONE(4), 
#COUNTY(5), COUNTYNAME(6), STATE(7), EVTYPE(8)

#Provide the column numbers
rawData.s <- select(rawData, 1:8)
str(rawData.s)
```

An alternative way...
```{r selectExample2, collapse=TRUE}
#Select only the following columns 
#BGN_DATE(2), BGN_TIME(3), TIME_ZONE(4), 
#COUNTYNAME(6), STATE(7), EVTYPE(8)

#Provide the column numbers
rawData.s <- select(rawData, c(2,3,4,6,7,8))
str(rawData.s)
```

###Select cols by col names  
```{r selectExample3, collapse=TRUE}
#Select only the following columns 
#STATE__(1), BGN_DATE(2), BGN_TIME(3), TIME_ZONE(4), 
#COUNTY(5), COUNTYNAME(6), STATE(7), EVTYPE(8)

#Provide the column names
rawData.s <- select(rawData, STATE__:EVTYPE)
str(rawData.s)
```

An alternative way...
```{r selectExample4, collapse=TRUE}
#Select only the following columns 
#STATE__(1), BGN_DATE(2) 
#EVTYPE(8)

#Provide the column names
rawData.s <- select(rawData, c(STATE__, BGN_DATE, EVTYPE))
str(rawData.s)
```

An alternative way...
```{r selectExample5, collapse=TRUE}
#Remove the following columns 
#STATE__(1), BGN_DATE(2) 
#EVTYPE(8)

#Provide the column names or column numbers
#-(minus) is used to tell to remove a specific element
rawData.s <- select(rawData, c(-STATE__, -BGN_DATE, -EVTYPE))
str(rawData.s)
```

###Select cols whose col names start with  
```{r selectExample6, collapse=TRUE}
#Select the columns whose names start with "BGN")

#starts_with is a special function
rawData.s <- select(rawData, starts_with("BGN"))
str(rawData.s)
```

###Select cols whose col names end with  
```{r selectExample7, collapse=TRUE}
#Select the columns whose names end with "DATE")

#ends_with is a special function
rawData.s <- select(rawData, ends_with("DATE"))
str(rawData.s)
```

##<a id="fil">`filter`verb</a>
The `filter` function is used to extract a subset of rows from a data frame based on a specific logic condition.

A simple filtering ....
```{r filterExample1, collapse = TRUE}
#Unique EVTYPE values
unique(rawData$EVTYPE)

#Original dimension raw dataset
dim(rawData)

#Getting all of the observation connecte with a "TORNADO" EVTYPE
rawData.f <- filter(rawData, EVTYPE == "TORNADO")

#Unique EVTYPE values
unique(rawData.f$EVTYPE)

#Dimension filtered dataset
dim(rawData.f)
```

A more complex filtering ....
```{r filterExample2, collapse = TRUE}
#Unique EVTYPE values
unique(rawData$EVTYPE)

#Unique COUNTYNAME values (first 10 ...)
unique(rawData$COUNTYNAME)[1:10]

#Original dimension raw dataset
dim(rawData)

#Getting all of the observation connecte with a "TORNADO" EVTYPE and "MOBILE" COUNTYNAME
rawData.f <- filter(rawData, EVTYPE == "TORNADO" & COUNTYNAME == "MOBILE")

#Unique EVTYPE values
unique(rawData.f$EVTYPE)

#Unique COUNTYNAME values
unique(rawData.f$COUNTYNAME)

#Dimension filtered dataset
dim(rawData.f)
```

##<a id="arr">`arrange`verb</a>
The `arrange` function is used to reorder rows of a data frame according to one of the col/ feature, preserving observations.

```{r collapse = TRUE}
head(rawData$TMP_BGN_DATE, 10)

tail(rawData$TMP_BGN_DATE, 10)
```
###Order ascending
Order the observations by `TMP_BGN_DATE` in ascending order (default).
```{r arrangeExample1, collapse = TRUE}
rawData.order.asc.by.TMP_BNG_DATE <- arrange(rawData, TMP_BGN_DATE)
```
Verify that observations have been actually ordered as expected.
```{r collapse = TRUE}
head(rawData.order.asc.by.TMP_BNG_DATE$TMP_BGN_DATE, 10)

tail(rawData.order.asc.by.TMP_BNG_DATE$TMP_BGN_DATE, 10)
```
###Order descending
Order the observations by `TMP_BGN_DATE` in descending order.
```{r arrangeExample2, collapse = TRUE}
rawData.order.desc.by.TMP_BNG_DATE <- arrange(rawData, desc(TMP_BGN_DATE))
```
Verify that observations have been actually ordered as expected.
```{r collapse = TRUE}
head(rawData.order.desc.by.TMP_BNG_DATE$TMP_BGN_DATE, 10)

tail(rawData.order.desc.by.TMP_BNG_DATE$TMP_BGN_DATE, 10)
```
##<a id="ren">`rename`verb</a>
The `rename` function is used to rename a variable in a data frame.

```{r collapse = TRUE}
#print out the names of the variables (cols)
names(rawData)[1:5]
```

```{r renameExample, collapse = TRUE}
rawData.r <- rename(rawData, STATE_ID = STATE__, BEGIN_DATE = BGN_DATE, BEGIN_TIME = BGN_TIME)
```

```{r collapse = TRUE}
#print out the names of the variables (cols)
names(rawData.r)[1:5]
```

##<a id="mut">`mutate`verb</a>
The `mutate` function is used to compute transformation of variables in a data frame. In other words to create new variables that are derived from existing variables.

Adding a new variable `TOTAL_VICTIMS` as the sum of `FATALITIES`and `INJURIES`.

```{r collapse = TRUE}
#print out the names of the variables (cols)
#check that TOTAL_VICTIMS does not exist.
names(rawData)
```

```{r mutateExample, collapse = TRUE}
rawData <- mutate(rawData, TOTAL_VICTIMS = FATALITIES + INJURIES)
```

```{r collapse = TRUE}
#print out some of the data
head(rawData[,c("FATALITIES", "INJURIES", "TOTAL_VICTIMS")], 10)
```

##<a id="gro">`group_by`verb</a>
The `group_by` function is used to generate summary statistics from the data frame within strata defined by a variable.

```{r collapse = TRUE}
rawData.group.by.TMP_BGN_YEAR <- group_by(rawData, TMP_BGN_YEAR)
#Note the object implementing the returned objected implemented group_by 
typeof(rawData.group.by.TMP_BGN_YEAR)
```

Then the `summarize` function is used to perform some statistical computation on the data structur (`list`) created by `group_by`.
```{r collpase = TRUE}
summarize(rawData.group.by.TMP_BGN_YEAR, 
          HEALTH_COST = mean(TOTAL_VICTIMS, na.rm = TRUE))
```